Supplementary material for: non-coding RNAs of bird genomes

### 2014-04-11 paper/manuscript.tex
Working on the RNA-seq validation.

{\bf needs work:} New sequencing technologies have dramatically
expanded the rate at which ncRNAs are discovered and their functions
are determined \cite{}. However, in order to determine the full range
of ncRNAs across multiple species we would require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive. Consequently, for this work we have concentrated on
bioinformatic approaches. Primarily, homology based methods, namely
covariance models (CMs). These remain state of the art for ncRNA
analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families. These are used to train
probabilistic models for each family. These models can be used to
generate sequences with similar properties, score the likelihood that
a sequence is generated by the same evolutionary processes as the
training sequences and to build alignments based upon sequence and
structural information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.
The tRNAscan-SE software package uses CMs to accurately predict
transfer RNAs \cite{Lowe:1997,Chan:2009}.  The Rfam database contains
thousands of curated alignments and consensus structures for diverse
classes of ncRNAs
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}.
Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
The CM based approach works well for almost all classes of ncRNA, but
the long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to
dramatic speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can
they deal simply with the repeats that many lncRNAs host. Consequently
in the latest release of Rfam the lncRNA families that were added were
composed of local conserved (and possibly structured elements) within
lncRNAs, analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. The functions determined to date for lncRNAs range
from regulating chromatin status to chromosomal inactivation
\cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
genes is a lengthy and expensive process \cite{Guttman:2009}.

@@

The ncRNAs presented here have been identified using homology models
and are evolutionarily conserved in multiple avian species. In order
to further validate these predictions we have used RNA-seq from
multiple chicken tissues. After mapping chicken RNA-seq data to the
chicken genome (see Methods for details), we identified a threshold
for calling a gene as expressed by limiting our estimated
false-positive rate to less than 10\%. This was estimated using
randomly selected, un-annotated and presumably un-expressed regions of
the genome. Some of which may be genuinely expressed, meaning the true
FDR is potentially much lower than 10\%. The $<10\%$ FDR constraint,
corresponded to identifying at least one experimental condition where
13 RNA-seq reads or more were identified, that mapped to a gene of
interest. Overall, the number of ncRNAs we have identified in this
work that are expressed above background levels is 68.4\% (see
Table~\ref{table:1}). This shows that 6.9 times more of our ncRNAs are
expressed than expected by chance (Fisher's exact test:
$P<10^{16}$). It is likely that this is an underestimate of the
fraction of our annotations that are genuinely expressed, as only a
fraction of the developmental stages and tissues essential for forming
a chicken have been sampled with RNA-seq. Furthermore we now know that
some ncRNAs are expressed in highly specific conditions \cite{} ({\bf
  the worm miRNA in 1 cell paper}).

The classes of RNAs where the majority of our annotations were
confirmed includes microRNAs, snoRNAs, cis-regulatory elements, tRNAs,
SRP RNA and RNase P/MRP RNA. The RNA-seq data did fail to detect a
telomerase RNA transcript ({\bf ARE THESE ONLY EXPRESSED DURING
  MEIOSIS?}). The fractions for the other classes were well above the
10\% FDR threshold but admittedly were low for 7SK RNA, the minor
spliceosomal RNAs and the lncRNAs. There are a number of possible
explanations for this: the multiple copies of the 7SK RNA may be
functionally redundant and can therefore compensate for one another;
The minor spliceosome is active during ({\bf SOME OBSCURE CONDITION
  HOPEFULLY???}); and the lncRNAs are generally expressed at low
levels under specific conditions ({\bf cite RNACAPTURE and MATTICK'S
  BRAIN ATLAS PAPER}).

@@

Our diverse sets of genome annotations were combined. We collapsed the
overlapping annotations into a single annotation. We also generated
heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:1}
and Sup. Figs. 1-3). All the scripts and annotations presented here
are available from Github \cite{gitrepo}.

{\bf Rewrite:}
Chicken snoRNA and miRNA annotations were validated using small
RNA-seq data comprising 27 samples from 14 diferent tissues. Since not
all RNAs are expressed at any given time we expect to verify only a
fraction of annotated RNAs using the expression data. Nevertheless 242
of 691 (35\%) miRNAs as well as 328 of 376 (87\%) snoRNAs of our
homology-based annotations were found to be supported by the RNA-seq
data.

{\bf WHERE IS ALL THE GENOME SEQUENCE AND RNA-SEQ DATA AVAILABLE
  FROM?}

@@

We thank Fiona McCarthy (Mississippi State University) and Carl
Schmidt (University of Delaware) for providing the RNA-seq data.

@@

We thank Matt Schwartz (Harvard) and Igor Ulitsky (Weizmann Institute
of Science) for access to the RNA-seq data.

@@

\begin{tabular}{|r|r|r|r|l|}
\hline 
\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
\hline 
Number in human & median(48 birds) & Number in chicken & Chicken ncRNAs         & RNA type\\
                &                  &                   & confirmed with RNA-seq & \\
\hline
62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
356&499.5&427&272 (63.7\%) &microRNA\\ 
281&120.0&106&88 (83.0\%) &C/D box snoRNA\\ 
336&85.5&68&46 (67.6\%) &H/ACA box snoRNA\\ 
34&13.0&12&11 (91.7\%) &Small cajal body RNA\\ 
1754&48.5&71&29 (40.8\%) &Major spliceosomal RNA\\ 
58&3.0&6&2 (33.3\%) &Minor spliceosomal RNA\\ 
525&82.0&122&79 (64.8\%) &Cis-regulatory element\\ 
316&6.5&9&3 (33.3\%) &7SK RNA\\ 
1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
9&0.0&2&1 (50.0\%) &Vault RNA\\ 
892&3.0&3&2 (66.7\%) &Y RNA\\ 
1084&173.5&300&256 (85.3\%) &Transfer RNA\\ 
80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
941&3.0&4&2 (50.0\%) &SRP RNA\\ 
607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
\hline
7340&1080.0&1194&817 (68.4\%) &Total\\ 
\hline
\end{tabular}

>>>>

New sequencing technologies have dramatically expanded the rate at
which ncRNAs are discovered and their functions are determined
\cite{cech2014noncoding}. However, in order to determine the full
range of ncRNAs across multiple species we would require multiple RNA
fractions (e.g. long and short), in multiple species, in multiple
developmental stages and tissues types. The costs of this approach are
still prohibitive. Consequently, for this work we have concentrated on
bioinformatic approaches. Primarily, homology based methods, namely
covariance models (CMs). These remain state of the art for ncRNA
analyses \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009} and have well
established sensitivity and specificity rates \cite{Freyhult:2007}.
The CM based approach for annotating ncRNAs in genomes requires
reliable alignments and consensus secondary structures of
representative sequences of RNA families, many of which can be found
at Rfam
\cite{Griffiths-Jones:2003,Griffiths-Jones:2005,Gardner:2009,Gardner:2011a,Burge:2013}. These
are used to train probabilistic models that score the likelihood that
a database sequence is generated by the same evolutionary processes as
the training sequences based upon both sequence and structural
information \cite{Sakakibara:1994,Eddy:1994,Nawrocki:2009}.  The
tRNAscan-SE software package uses CMs to accurately predict transfer
RNAs \cite{Lowe:1997,Chan:2009}.

Independent benchmarks of bioinformatic annotation tools have shown
that the CM approaches dramatically out-perform alternative methods
\cite{Freyhult:2007}, although its sensitivity is limited for the most
rapidly evolving families such as vault RNAs or telomerase RNA
\cite{Menzel:09a}.

%%{\bf REDUCE? too much talking about what it can't to compared to what it can do...}
The CM based approach works well for most classes of ncRNA, but the
long non-coding RNAs (lncRNAs) are a particular challenge
\cite{Guttman:2009}. Recent technological advances have led to
dramatic speed and memory-usage enhancements for CM analyses
\cite{Eddy:2002,Nawrocki:2007,Nawrocki:2009,Eddy:2011}. However, CMs
cannot model the exon-intron structures of spliced lncRNAs, nor can
they deal simply with the repeats that many lncRNAs host. Consequently
in the latest release of Rfam the lncRNA families that were added were
composed of local conserved (and possibly structured elements) within
lncRNAs, analogous to the ``domains'' housed within protein sequences
\cite{Burge:2013}. Whilst some these regions may not reflect
functional RNA elements instead regulatory regions, enhancers or
insulators, their syntenic conservation still provides an indication
of lncRNA conservation \cite{diederichs2014four}.

%% The functions determined to date for lncRNAs range
%% from regulating chromatin status to chromosomal inactivation
%% \cite{Rinn:2007,Chow:2005}. Yet functional characterisation of these
%% genes is a lengthy and expensive process \cite{Guttman:2009}.

@@

The ncRNAs presented here have been identified using homology models
and are evolutionarily conserved in multiple avian species. In order
to further validate these predictions we have used strand-specific
total RNA-seq and small RNA-seq of multiple chicken tissues. After
mapping the RNA-seq data to the chicken genome (see Methods for
details), we identified a threshold for calling a gene as
expressed by limiting our estimated false-positive rate to less than
10\%. This FDR was estimated using a negative control of randomly
selected, un-annotated regions of the genome. Since some regions may
be genuinely expressed, the true FDR is potentially much lower than
10\%. Overall, the number of ncRNAs we
have identified in this work that are expressed above background
levels is 865 (72.4\%) (see Table~\ref{table:1}). This shows that 7.0
times more of our ncRNAs are expressed than expected by chance
(Fisher's exact test: $P<10^{16}$). This number is an underestimate of
the fraction of our annotations that are genuinely expressed, as only
a fraction of the developmental stages and tissues of chicken have
been characterized with RNA-seq. Furthermore, some ncRNAs
are expressed in highly specific conditions  \cite{mercer2008specific,johnston2003microrna}.

The classes of RNAs where the majority of our annotations were
experimentally confirmed includes microRNAs, snoRNAs, cis-regulatory
elements, tRNAs, SRP RNA and RNase P/MRP RNA. The RNA-seq data could
not provide evidence for a telomerase RNA transcript, which are only
generally only expressed in embryonic, stem or cancerous tissues. Only
a small fraction of the 7SK RNA, the minor spliceosomal RNAs and the
lncRNAs could be confirmed with the 10\% FDR threshold. There are a
number of possible explanations for this: the multiple copies of the
7SK RNA may be functionally redundant and can therefore compensate for
one another; The minor spliceosome is, as the name suggests, a rarely
used alternative spliceosome; and the lncRNAs are generally expressed
at low levels under specific conditions
\cite{mercer2008specific,mercer2012targeted}.

@@

The diverse sets of genome annotations were combined. We collapsed the
overlapping annotations into a single annotation. We also generated
heatmaps for different groups of ncRNA genes (see Figure~\ref{fig:1}
and Sup. Figs. 1-3). All the scripts and annotations presented here
are available from Github \cite{gitrepo}.

Chicken ncRNA predictions were validated using two separate RNA-seq
data sets. The first data set (Bioproject PRJNA204941) contains 971
million reads and comprises 27 samples from 14 diferent chicken
tissues sequenced on Illumina HiSeq2000 using a small RNA-seq
protocol. The second data set (personal commun. Matt Schwartz and Igor
Ulitsky) contains 1,46 billion Illumina HiSeq reads sequenced from
whole chicken embryo RNA from 7 stages using a strand-specific dUTP
protocol. Raw reads were checked for quality and adapters clipped if
required by the protocol. Preprocessed reads were mapped to the
galGal4 reference genome using SEGEMEHL short read aligner
\cite{segemehl} and then overlapped with the ncRNA annotations.

@@

We thank Fiona McCarthy (Mississippi State University) and Carl
Schmidt (University of Delaware) as well as Matt Schwartz (Harvard)
and Igor Ulitsky (Weizmann Institute of Science) for providing the
RNA-seq data as part of the Avian RNAseq consortium.

@@

@@

%% \begin{tabular}{|r|r|r|r|l|}
%% \hline 
%% \multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
%% \hline 
%% Number in human & median(48 birds) & Number in chicken & Chicken ncRNAs         & RNA type\\
%%                 &                  &                   & confirmed with RNA-seq & \\
%% \hline
%% 62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
%% 356&499.5&427&272 (63.7\%) &microRNA\\ 
%% 281&120.0&106&88 (83.0\%) &C/D box snoRNA\\ 
%% 336&85.5&68&46 (67.6\%) &H/ACA box snoRNA\\ 
%% 34&13.0&12&11 (91.7\%) &Small cajal body RNA\\ 
%% 1754&48.5&71&29 (40.8\%) &Major spliceosomal RNA\\ 
%% 58&3.0&6&2 (33.3\%) &Minor spliceosomal RNA\\ 
%% 525&82.0&122&79 (64.8\%) &Cis-regulatory element\\ 
%% 316&6.5&9&3 (33.3\%) &7SK RNA\\ 
%% 1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
%% 9&0.0&2&1 (50.0\%) &Vault RNA\\ 
%% 892&3.0&3&2 (66.7\%) &Y RNA\\ 
%% 1084&173.5&300&256 (85.3\%) &Transfer RNA\\ 
%% 80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
%% 941&3.0&4&2 (50.0\%) &SRP RNA\\ 
%% 607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
%% 4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
%% \hline
%% 7340&1080.0&1194&817 (68.4\%) &Total\\ 
%% \hline
%% \end{tabular}

\begin{tabular}{|r|r|r|r|l|}
\hline 
\multicolumn{5}{|l|}{{\bf ncRNA genes in human, chicken and all bird genomes}}\\
\hline 
                &                  &                   & Chicken ncRNAs & \\
Number in human & median(48 birds) & Number in chicken & confirmed with RNA-seq & RNA type\\
\hline
62&25.0&34&12 (35.3\%) &Long non-coding RNA\\ 
356&499.5&427&280 (65.6\%) &microRNA\\ 
281&120.0&106&90 (84.9\%) &C/D box snoRNA\\ 
336&85.5&68&48 (70.6\%) &H/ACA box snoRNA\\ 
34&13.0&12&12 (100.0\%) &Small cajal body RNA\\ 
1754&48.5&71&32 (45.1\%) &Major spliceosomal RNA\\ 
58&3.0&6&3 (50.0\%) &Minor spliceosomal RNA\\ 
525&82.0&122&88 (72.1\%) &Cis-regulatory element\\ 
316&6.5&9&3 (33.3\%) &7SK RNA\\ 
1&0.0&2&0 (0.0\%) &Telomerase RNA\\ 
9&0.0&2&1 (50.0\%) &Vault RNA\\ 
892&3.0&3&2 (66.7\%) &Y RNA\\ 
1084&173.5&300&278 (92.7\%) &Transfer RNA\\ 
80&9.5&4&2 (50.0\%) &Transfer RNA pseudogene\\ 
941&3.0&4&2 (50.0\%) &SRP RNA\\ 
607&7.0&22&10 (45.5\%) &Ribosomal RNA\\ 
4&1.0&2&2 (100.0\%) &RNase P/MRP RNA\\ 
\hline
7340&1080.0&1194&865 (72.4\%) &Total\\ 
\hline
\end{tabular}

%totalExpressed[865] totalNegControlsExpressed[123] totalGenes[1194] FPR[10.3]

### 2014-04-11 paper/supplementary-material.tex
Working on the RNA-seq validation.

>>>>

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr1-147253243-147253923-miRNA-cluster.pdf}
  \caption[]{microRNA cluster: mir-19\_3, mir-20\_2, mir-19\_2, mir-18\_2 \& mir-17\_1.}\label{fig:10}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr4-3946041-3946705-miRNA-cluster.pdf}
  \caption[]{microRNA cluster: mir-92\_2, mir-19\_1, mir-20\_1, mir-18\_1, mir-106\_1.}\label{fig:11}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr23-2507395-2508461-snoRNA-cluster.pdf}
  \caption[]{H/ACA box snoRNA cluster: SNORA61, SNORA44, SNORA16.}\label{fig:12}
\end{figure}

\begin{figure}[ht]
\includegraphics[width=0.99\textwidth]{/home/ppg15/people/birds/mario/chr27-4040511-4042052-tRNA-cluster.pdf}
  \caption[]{Cysteine tRNA cluster.}\label{fig:13}
\end{figure}
